---
title: "HOW TO CALCULATE CLIMATOLOGY FROM A GRIDDED DATASET"
author: "Fernando Prudencio"
date: "August 17, 2020"
output: github_document
---
## Load libraries
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(Hmisc)
library(raster)
library(ncdf4)
library(tmap)
library(cptcity)
```
## Create time series within a dataframe
Before calculating the **climatology** of any variable, for a given period (eg: **1981-2016**), first a time series field is created within a dataframe, in addition to an identifier field **[id]**.

Build a data frame with a date field and an identity field for the period from 1981 to 2016 (by month)
```{r message = FALSE, warning = FALSE}
df <- tibble(
  date = seq(as.Date("1981-01-01"), as.Date("2016-12-01"), by = "1 month")
) %>%
  mutate(id = 1:n())
```
```{r, echo = FALSE}
head(df)
```
## Build a function to calculate climatology
In this step, we are going to build a function to calculate climatology without considering years of extreme drought
```{r echo=TRUE, message=FALSE, warning=FALSE}
fun.clim <- function(month, years.omit, data) {
  grd.mt <- df %>%
    filter(
      str_sub(date, 6, 7) == month &
        str_sub(date, 1, 4) %nin% years.omit
    )

  data[[grd.mt$id]] %>%
    "*"(1) %>%
    mean(na.rm = T) %>%
    return()
}
```
## Apply fun.clim function with sapply()
Apply **fun.clim** function to calculate average conditions between january and december. Omitted years were 2005, 2010 and 2016, as mentioned in ***Marengo and Espinoza (2016)***. For this example, we going to use the Pisco v2.1 gridded precipitation dataset.
```{r echo=TRUE, message=FALSE, warning=FALSE}
grd.clim <- sapply(
  sprintf("%02d", 1:12),
  FUN = fun.clim,
  years.omit = c(2005, 2010, 2016),
  data = brick("data/raster/pp/pisco-pp.nc")
) %>% stack() %>% "*"(1) %>%
  "names<-"(sprintf(month.abb))
```
```{r echo = FALSE}
grd.clim
```
## Save climatology raster as .grd format
Save the climatology raster as **.grd** format to keep the name of the bands (also see **netCDF** format)
```{r include = TRUE}
writeRaster(grd.clim, "data/raster/pisco_clim.grd", overwrite = T)
```
## Plot average conditions for February
In this step, we going to use the tmap package and the **World** vector
```{r message=FALSE, warning=FALSE}
data("World")
tmap_mode("plot")
tm_shape(grd.clim[["Feb"]]) +
  tm_raster(
    title = "pp [mm]",
    style = "kmeans", n = 5,
    palette = cpt(
      pal = "ncl_precip_diff_12lev", n = 100,
      colorRampPalette = FALSE, rev = T
    ),
    legend.hist = TRUE
  ) +
  tm_shape(World, bbox = World %>% filter(name == "Peru")) +
  tm_borders(col = "black") +
  tm_legend(outside = TRUE, hist.width = 2) +
  tm_grid(
    labels.inside.frame = FALSE,
    x = seq(-85, -65, 5),
    y = seq(-20, 5, 5),
    projection = "+proj=longlat"
  ) +
  tm_layout(bg.color = "gray")
```